\textbf{Varying coefficients correlated velocity models in complex landscapes with application to narwhal behavioral response to ship exposure and airgun exposure}

### 2024-05-29 Latex/main.tex
updates

%%PACKAGES
\usepackage{authblk}
\usepackage{hyperref} 
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{multirow}

@@

 	\caption{Five day trajectories with time step $\Delta=5$ min within circular, rectangular and fjords domains. Plot labeled persistent and standard have the same smooth parameter $\omega$ defined in equation~\ref{eq: smooth omega}, but the persistent trajectory as a smooth parameter $\tau$ as in Figure~\ref{fig:smoothtaubump} while the standard trajectory has a constant $\tau =1$h. Trajectories labeled "tortuous" are obtained from a contant $\tau=1$ h and a smooth parameter $\nu$ defined with spline basis function as in figures~\ref{fig: smoothomegasplines1}~\ref{fig: smoothomegasplines2}.}
 	\label{fig: crcvm examples}
 \end{figure}

@@

\section{Simulation study for the baseline model}
\label{section: simulation constrained motion}
In order to experiment the constrained models we simulate a CRCVM in Scoresby Sound fjords system in which the narwhals move in our application. We consider observations of a baseline model as described in equation~\ref{eq: exact baseline model}. We set $L=9$ degrees of freedom for the bivariates splines in the parameter $\omega$ and  we place the knots regularly between $-\pi$ and $\pi$ for the covariate $\Theta$, and between $0.5$ and $5$ km for $D^{shore}$. The values of the spline coefficients are chosen before hand to get a process that meet the spatial constraints. We fix $\tau_0=1$ h and $\nu_0=4$ km/h. Random effects standard deviations for the parameters $\tau$ and $\nu$ are set to $\sigma_{\tau}=0.2$ and $\sigma_{\nu}=0.1$. Initial velocities are set to $(0,0)$ and initial positions are uniformly sampled at least $1$ km away from the shore in Hjornedal region where the narwhals were initially captured and released for the controlled exposure experiment. \\

@@

None of the $600$ simulations reached the boundary. \\

@@

These $100$ batches of noisy samples are used as data for estimation. First, we estimate using only the first $12$h of the trajectories, that is $720$ observations for each narwhals. Then, we use the whole $24$h-long trajectory. The covariates $E^{shore}$ and $\Theta$ are computed from the observed noisy positions. It is worth noting that as measurement error grows, they are becoming less accurate, which makes estimation more challenging. The smooth parameter $\omega$ is estimated using bivariate tensor splines of these covariates, given by the function \texttt{te} in \texttt{R}. The marginal spline degree of freedom is set to $3$ for each covariate to match the fixed degree of freedom $L=8$. This corresponds to one single knot in each dimension. Hence, $9$ coefficients, including the intercept, are to be estimated for the smooth function $\omega$, plus the two parameters $\log(\tau_0)$ and $\log(\nu_0)$, and the individual effects standard deviations $\sigma_{\tau}$ and $\sigma_{\nu}$. Measurement error $\sigma_{obs}$ is supposed to be known.  
We fix initial parameter values to $\tau_{0}=1$ h, $\nu_{0} = 1$ km/h, $\sigma_{\tau}=1$, $\sigma_{\nu}=1$ and $\omega_k=0$ for all $k \in \{0,\cdots,L\}$. 
For each batch of $N=6$ trajectories, we obtain estimates $\widehat{\log(\tau_{0})}^{(k)}$, $\widehat{\log(\nu_{0})}^{(k)}$, $\hat{\omega}_l^{(k)}, l \in \{0,\cdots, L\}$, $\hat{\sigma}_{\tau}^{(k)}$ and $\hat{\sigma}_{\nu}^{(k)}$, $k \in \{1,\cdots,M\}$ and compute the root mean squared error and the biais of the estimates. Simulation and optimization of the log-likelihood for one batch of trajectories is performed on $16$ cpu cores within less than $1$ h. The results are summarized in Tables~\ref{table: estimation performance fjords 12h} and~\ref{table: estimation performance fjords 24h}.
Figure~\ref{fig: marginal_estimates_CRCVM} shows the results of the estimation of $f_{\omega}$ from one batch of $6$ trajectories with confidence intervals obtained from the observed Fisher information matrix. The function $f_{\omega}$  is plotted as a function of $\Theta$ for different fixed values of the distance to the shore, representing movement relatively close and far from the shore.

@@

\begin{table}[H]
	\centering
		\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|c|}
		\hline
		&\multicolumn{3}{|c|}{Intercepts} & \multicolumn{8}{|c|}{Spline coefficients} & \multicolumn{2}{|c|}{SD} \\
		\hline
		&$\log(\tau_0)$ & $\log(\nu_0)$ & $\omega_0$ & $\omega_1$ & $\omega_2$ & $\omega_3$ & $\omega_4$  & $\omega_5$ & $\omega_6$ & $\omega_7$ & $\omega_8$ & $\sigma_{\tau}$ & $\sigma_{\nu}$ \\
		\hline 
		 True values & $0$ & $1.39$ & $0.01$ & $6.47$ & $-2.07$ & $4.62$ & $15.44$ & $4.62$ & $21.10$ & $2.91$ & $4.86$ & $0.2$ & $0.1$ \\
		\hline
			\multicolumn{14}{|c|}{High frequency} \\
			\hline
	\multirow{2}{4em}{$N=6$} & & & & & & & & & & & & &\\
	& & & & & & & & & & & & & \\
	\hline
	\multirow{2}{4em}{$N=12$} & -0.03 & 1.38 & -0.02
	&5.82&-1.60 & 3.84 & 14.59 &4.06 & 19.77 & 3.04 & 4.20 & 0.09& 
	0.09\\
	& 0.09 & 0.04 & 0.17 & 0.64 &0.83 &0.81 & 1.08 & 1.83 & 1.02 &0.65 &0.77 &
	0.08 & 0.03  \\
	\hline	
	\multicolumn{14}{|c|}{Low frequency} \\
	\hline
	\multirow{2}{4em}{$N=6$} & & & & & & & & & & & & &\\
	& & & & & & & & & & & & & \\
	\hline
	\multirow{2}{4em}{$N=12$} & -0.03 & 1.38 & -0.02
	&5.82&-1.60 & 3.84 & 14.59 &4.06 & 19.77 & 3.04 & 4.20 & 0.09& 
	0.09\\
	& 0.09 & 0.04 & 0.17 & 0.64 &0.83 &0.81 & 1.08 & 1.83 & 1.02 &0.65 &0.77 &
	0.08 & 0.03  \\
	\hline	
\end{tabular}
\end{table}

>>>>

%%PACKAGES
\usepackage{authblk}
\usepackage{hyperref} 
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{multirow}
\usepackage{lscape}

@@

 	\caption{Five day trajectories with time step $\Delta=5$ min within circular, rectangular and fjords domains. Plots labeled "persistent" and "standard" have the same smooth parameter $\omega$ defined in equation~\ref{eq: smooth omega}, but the persistent trajectory as a smooth parameter $\tau$ as in Figure~\ref{fig:smoothtaubump} while the standard trajectory has a constant $\tau =1$h. Trajectories labeled "tortuous" are obtained from a constant $\tau=1$ h and a smooth parameter $\nu$ defined with spline basis function as in figures~\ref{fig: smoothomegasplines1}~\ref{fig: smoothomegasplines2}.}
 	\label{fig: crcvm examples}
 \end{figure}

@@

\section{Simulation study for the baseline model}
\label{section: simulation constrained motion}
In order to experiment estimation of the constrained models we simulate CRCVM trajectories in Scoresby Sound fjords system where the narwhals move in our application. We consider observations of a baseline model as described in equation~\ref{eq: exact baseline model}. We set $L=9$ degrees of freedom for the bivariates splines in the parameter $\omega$ and  we place the knots regularly between $-\pi$ and $\pi$ for the covariate $\Theta$, and between $0.5$ and $5$ km for $D^{shore}$. The values of the spline coefficients are chosen before hand to get a process that meet the spatial constraints. We fix $\tau_0=1$ h and $\nu_0=4$ km/h. Random effects standard deviations for the parameters $\tau$ and $\nu$ are set to $\sigma_{\tau}=0.2$ and $\sigma_{\nu}=0.1$. Initial velocities are set to $(0,0)$ and initial positions are uniformly sampled at least $1$ km away from the shore in Hjornedal region where the narwhals were initially captured and released during the controlled exposure experiment. \\

@@

A small percentage (around $2\%$) of all the simulated trajectories reached the shoreline, due to regions of acute corner in the polygon defining land. For the estimation, we only kept the trajectories that did not reached land and whose observed distance to the shore in km were included in the interval $[0.5,5]$ for consistency with the fixed knots used to define the true smooth function $\omega$.  \\

@@

These batches of noisy samples are used as data for estimation. It is worth noting that as measurement error grows, they are becoming less accurate, which makes estimation more challenging in the low frequency framework. For the smooth $\omega$, $9$ coefficients, including the intercept, are to be estimated, plus the two parameters $\log(\tau_0)$ and $\log(\nu_0)$, and the individual effects standard deviations $\sigma_{\tau}$ and $\sigma_{\nu}$. Measurement error $\sigma_{obs}$ is supposed to be known.  
We fix initial parameter values to $\tau_{0}=1$ h, $\nu_{0} = 1$ km/h, $\sigma_{\tau}=1$, $\sigma_{\nu}=1$ and $\omega_k=0$ for all $k \in \{0,\cdots,L\}$. 
For each batch of $N$ trajectories, we obtain estimates $\widehat{\log(\tau_{0})}^{(k)}$, $\widehat{\log(\nu_{0})}^{(k)}$, $\hat{\omega}_l^{(k)}, l \in \{0,\cdots, L\}$, $\hat{\sigma}_{\tau}^{(k)}$ and $\hat{\sigma}_{\nu}^{(k)}$, $k \in \{1,\cdots,M\}$ and compute the mean  and the standard deviation of the estimates. Simulation and optimization of the log-likelihood for one batch of trajectories is performed on $16$ cpu cores within less than $1$ h. The results are summarized in Table~\ref{table: simulation study results}.

\begin{landscape}
\begin{table}[H]
	\centering
		\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|c|c|}
		\hline
		&\multicolumn{3}{|c|}{Intercepts} & \multicolumn{8}{|c|}{Spline coefficients} & \multicolumn{2}{|c|}{SD} \\
		\hline
		&$\log(\tau_0)$ & $\log(\nu_0)$ & $\omega_0$ & $\omega_1$ & $\omega_2$ & $\omega_3$ & $\omega_4$  & $\omega_5$ & $\omega_6$ & $\omega_7$ & $\omega_8$ & $\sigma_{\tau}$ & $\sigma_{\nu}$ \\
		\hline 
		 True & $0$ & $1.39$ & $0.01$ & $6.47$ & $-2.07$ & $4.62$ & $15.44$ & $4.62$ & $21.10$ & $2.91$ & $4.86$ & $0.2$ & $0.1$ \\
		\hline
			\multicolumn{14}{|c|}{High frequency} \\
			\hline
	\multirow{2}{4em}{$N=6$} & & & & & & & & & & & & &\\
	& & & & & & & & & & & & & \\
	\hline
	\multirow{2}{4em}{$N=12$} & -0.03 & 1.38 & -0.02
	&5.82&-1.60 & 3.84 & 14.59 &4.06 & 19.77 & 3.04 & 4.20 & 0.09& 
	0.09\\
	& $\pm$ 0.09 & $\pm$ 0.04 & $\pm$ 0.17 & $\pm$ 0.64 & $\pm$ 0.83 & $\pm$ 0.81 & $\pm$ 1.08 & $\pm$ 1.83 & $\pm$ 1.02 & $\pm$ 0.65 & $\pm$ 0.77 &
	$\pm$ 0.08 & $\pm$ 0.03  \\
	\hline	
	\multicolumn{14}{|c|}{Low frequency} \\
	\hline
	\multirow{2}{4em}{$N=6$} & & & & & & & & & & & & &\\
	& & & & & & & & & & & & & \\
	\hline
	\multirow{2}{4em}{$N=12$} & -0.03 & 1.38 & -0.02
	&5.82&-1.60 & 3.84 & 14.59 &4.06 & 19.77 & 3.04 & 4.20 & 0.09& 
	0.09\\
	& 0.09 & 0.04 & 0.17 & 0.64 &0.83 &0.81 & 1.08 & 1.83 & 1.02 &0.65 &0.77 &
	0.08 & 0.03  \\
	\hline	
\end{tabular}
\label{table: simulation study results}
\caption{Simulation study results}
\end{table}
\end{landscape}

In practical case, confidence intervals on the estimates are derived from the observed Fisher information matrix. Figure~\ref{fig: marginal_estimates_CRCVM} shows the results of the estimation of $f_{\omega}$ from one batch of $6$ trajectories with these confidence intervals.  The function $f_{\omega}$  is plotted as a function of $\Theta$ for different fixed values of the distance to the shore, representing movement relatively close and far from the shore.

@@
